\section{Evaluation}
\label{sec:eval}

\TODO{Draft after all experiments run. Outline below reflects the locked study structure.}

\subsection{Experimental Setup}
\TODO{hardware, PG16 config, agent models, temperatures, query caps, benchmark datasets (IMDB/JOB and curated InsightBench subset)}

\subsection{E1: End-to-End Headline on InsightBench}
\TODO{Primary headline result. Protocol vs.\ Baseline A across \placeholder{} curated tasks. Table of per-task session time and quality-rubric agreement.}

\subsection{E2: End-to-End Mechanism Study on IMDB}
\TODO{Per-scenario session time breakdown across 20 reps. Table below is the structure; numbers update when pre-hoc and M2 land.}

%% Placeholder table showing the result structure; numbers from current post-hoc M1 data
\begin{table}[h]
\centering
\small
\caption{IMDB session-time breakdown, M1 post-hoc implementation (20 reps across 10 scenarios). \TODO{Replace with pre-hoc + M2 + M3 numbers when measured.}}
\label{tab:e2-imdb}
\begin{tabular}{lrrrr}
\toprule
Scenario & With M1 (s) & Baseline (s) & Savings (s) & Speedup \\
\midrule
1.\ Genre Evolution & 29.8 & 51.1 & 21.3 & 1.72$\times$ \\
2.\ Director Career & 393.5 & 496.8 & 103.3 & 1.26$\times$ \\
3.\ Company Genre & 34.3 & 43.2 & 8.9 & 1.26$\times$ \\
4.\ Cast Size & 100.7 & 117.0 & 16.4 & 1.16$\times$ \\
5.\ Co-Production & 36.9 & 51.1 & 14.2 & 1.38$\times$ \\
6.\ Franchise & 10.9 & 15.5 & 4.5 & 1.41$\times$ \\
7.\ Writer-Director & 18.7 & 24.7 & 6.0 & 1.32$\times$ \\
8.\ Actor Archetypes & 198.9 & 319.4 & 120.5 & 1.61$\times$ \\
9.\ Spillover & 235.1 & 256.0 & 20.9 & 1.09$\times$ \\
10.\ Budget & 16.6 & 23.4 & 6.8 & 1.41$\times$ \\
\midrule
\textbf{TOTAL} & \textbf{1075.3} & \textbf{1398.1} & \textbf{322.7} & \textbf{\imdbSpeedupPostHoc} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{E3: Per-Mechanism Ablation}
\TODO{M1-only / M1+M2 / M1+M2+M3 / full protocol on IMDB. Bar chart or stacked bar.}

\subsection{E4: Agent Adoption and Declaration Faithfulness}
\TODO{Voluntary adoption rates, SAVE\_CTE distribution, M2 precision; documented failure mode ``iteration on base''}

Observed across the 20-rep IMDB suite under M1 post-hoc:
\begin{itemize}
\item M1 adoption: \adoptionRate voluntary save rate (15/20 reps).
\item SAVE\_CTE dominance: \saveCteRate of reps use SAVE\_CTE specifically.
\item M2 declaration rate: \mTwoDeclRate (9/20 reps).
\item M2 precision: \mTwoPrecision of declared variants actually issued.
\item Task completion: \taskCompletionIMDB.
\end{itemize}

\subsection{E5: Intra- vs.\ Inter-Query Reuse}
\TODO{Single-ref vs.\ multi-ref CTE subset. Measures Concord's contribution beyond PG16's built-in multi-ref CTE materialization.}

Current post-hoc numbers:
\begin{itemize}
\item Single-reference reuse subset (44 queries): \imdbSingleRefSpeedup speedup. PG16 would inline the CTE; Concord's materialization is where the win comes from.
\item Multi-reference subset (25 queries): \imdbMultiRefSpeedup speedup. PG16 already materializes; Concord's delta is smaller (but still positive, reflecting physical-property benefits from M3).
\end{itemize}

\subsection{E6: Sensitivity}
\TODO{Lookahead depth $k \in \{1,2,3,5\}$; agent capability axis (Opus / Haiku / mid-tier); data scale (IMDB 10\% / 100\%).}

\subsection{Quality Evaluation}
\TODO{Pre-registered rubric, blinded human grading of paired outputs, LLM-judge cross-validation, symmetric reporting (agreement / protocol-wins / baseline-wins rates).}
